---
title: "Hierarchical Models or Mixed Models"
author: "Amanda Franklin"
date: "04/04/2022"
output: html_document
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../docs") })
---

## Introduction

This code conducts a generalised linear mixed model, or a hierarchical model
following recommendations in Zuur et al. 2009, Mixed Effects Models and 
Extensions in Ecology with R. 

Zuur recommends the following "cookbook" approach
1. Start with a model with all random and all fixed effects
2. Use LRT/AIC/BIC to decide which random effects to keep
3. Re-run winning model with REML = F (i.e. use ML rather than REML)
4. Use marginal hypothesis tests to evaluate fixed effects

If you want to know why all these steps, read Chapter 5 in Zuur

## Libraries 

```{r message = F, error = F}
library(lme4) # lmer()
library(tidyverse) # data processing functions and piping
library(bbmle) # ICtabs()
library(car) # Anova()

```

## 1. Dataset information

Grasshopper example

This dataset has:
- 9 roadside populations and 9 non-roadside populations
- 5-10 grasshoppers per population
- 3 replicate recordings per grasshopper
- quiet or loud rearing environments

LocMax is the maximum calling frequency (Hz)

```{r import data}
dat_gh <- read.csv("../data/GLMM.csv")
head(dat_gh)

```

## 2. Look at Variation 

This section explains the hierarchical nature of the dataset and demonstrates
the sources of variation in the dataset. You may skip this section if you
understand this.

First let's keep it simple and look at controls (quiet, non-roadside)
Where is the variation? We have:
1. Variation among populations
2. Variation among individuals
3. Variation among songs within individuals

We can look at the variation at each level
Select controls only

```{r}
dat_gh1 <- dat_gh %>%
  filter(Origin == "non-roadside", 
         Treatment == "quiet")
```

#### Observed variance within individuals
```{r}
varSvec <-  tapply(dat_gh1$LocMax, dat_gh1$Individual, var, na.rm = T) # vector of variances among songs within individuals
varS <-  mean(varSvec, na.rm = T)
varS
```


#### Observed variance among individual average values
```{r}
meanInd <- tapply(dat_gh1$LocMax, dat_gh1$Individual, mean, na.rm = T) # means for each individual
pops <- substr(names(meanInd), start = 1, stop = 2) # identifier for individuals by population
varIvec <- tapply(meanInd, pops, var, na.rm = T) # vector of variances among individual means
varI <- mean(varIvec, na.rm = T)
varI
```

#### Observed variance among population average values
```{r}
meanPop <-  tapply(meanInd, pops, mean, na.rm = T) # means for each individual
varP <-  var(meanPop)
varP
```

## 3. Run GLMM 

Let's run a GLMM following the Zuur guidelines

Does inclusion of random effects change our conclusions about effects of noise 
on songs?

#### Model with no random effects
```{r}
m1a <-  glm(LocMax ~ Origin*Treatment, data = dat_gh)
Anova(m1a)
```

#### Full model (i.e. Zuur step 1)
```{r}
m1b <- lmer(LocMax ~ Origin*Treatment + (1|Individual) + (1|Site), 
            data = dat_gh)
Anova(m1b)
```

Note: the p-values are quite different, the random effects can have a 
substantial influence on the p-values (usually increasing)

#### Look at the output:
```{r}
summary(m1b)

```
Random effect section shows the variation among individuals, sites and among 
songs within individuals (residual)


#### What random effects should we include? (Zuur Step 2)
We can use likelihood ratio tests, AIC or BIC to decide. 
Here we use AIC.

```{r}
m2a <- glm(LocMax ~ Origin*Treatment, data = dat_gh)
m2b <- lmer(LocMax ~ Origin*Treatment + (1|Individual) + (1|Site), 
            data = dat_gh)
m2c <- lmer(LocMax ~ Origin*Treatment + (1|Individual), data = dat_gh)
m2d <- lmer(LocMax ~ Origin*Treatment + (1|Site), data = dat_gh)

ICtab(m2a, m2b, m2c, m2d)
```

This indicates c and b are best. Since c and b are similar, and d is worse, 
site doesn't really improve the fit. Only Individual needs to be included as
a random effect

#### Run best model but using ML (i.e. REML = F, Zuur Step 3)
```{r}
m2c.1 <- lmer(LocMax ~ Origin*Treatment + (1|Individual), 
              data = dat_gh, REML = F)
```


#### Use marginal hypothesis tests to evaluate fixed effects (Zuur step 4)
```{r}
Anova(m2c.1)
Anova(m2c)
```

Results are similar for ML and REML
Strong evidence for effect of Origin, less for effect of Treatment


## 4. Plot 

#### Create model to get MLEs for plotting

```{r}
m2c.2 <- lmer(LocMax ~ -1 + Origin:Treatment + (1|Individual), 
              data = dat_gh, REML = F)
summary(m2c.2)
LocMax <- fixef(m2c.2)
group.CIs <- confint(m2c.2)[3:6,]
colnames(group.CIs) <- c("lower", "upper")
Origin <- c("Non-road", "Road", "Non-road", "Road")
`Rearing Condition` <- c("Noisy", "Noisy", "Quiet", "Quiet")
dat_plot <- data.frame(Origin, `Rearing Condition`, LocMax, group.CIs)

# If you want, you can also plot the raw data
# We'll make some changes so the datasets match
dat_gh2 <- dat_gh %>%
  mutate(Origin = ifelse(Origin == "roadside", "Road", "Non-road"),
         Treatment = ifelse(Treatment == "noisy", "Noisy", "Quiet")) %>%
  dplyr::rename(`Rearing Condition` = Treatment)

plot1 <- ggplot(dat_gh2, aes(x = Origin, y = LocMax, 
                             group = `Rearing Condition`)) + 
  geom_point(aes(colour = `Rearing Condition`), 
             position = position_jitterdodge()) +
  geom_point(data = dat_plot, aes(x = Origin, y = LocMax),
             position = position_dodge(width = 0.9), colour = "black") + 
  geom_errorbar(data = dat_plot, aes(ymin = lower, ymax = upper),
                position = position_dodge(width = 0.9)) + 
  theme_classic() + 
  ylab("Local Max Song Freq (Hz)") + xlab("Origin of Grasshopper")
plot1
```
